A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-28 response/response.tex

{\color{red}
  \begin{itemize}
    \item 
      The added subsection 'Convergence and displacement accuracy studies'
      includes convergence rate and displacement accuracy studies similar to
      those shown in 'Working examples ...', but with the models and tools of
      the main study. This subsection provides an example of interest. 
  \item 
      To summarize, these studies are done in \emph{Peridigm} using the most refined mesh
      from the 'serial' test runs from the original manuscript. Measured was the performance of the methods in terms of speed
      (convergence rate) and accuracy (displacement accuracy).  Iteration speed
      was reported in the original manuscript.
  \item
      It was determined that the accuracy of CS over FD resulted in no appreciable
      advantage in terms of displacement accuracy or convergence rate for the problems
      solved in this study. 
    \item
      Additional comment on the significance of the accuracy of the Jacobian
      to Newton's method is included in the subsections 'For further studies of
      Jacobian accuracy in Newton's method', 'Grain of salt for speed results'
      and 'Remarks'. 
  \item 
      To summarize, Newton's method is insensitive to the accuracy differences
      shown by the methods, for the problems shown, when the methods are not
      deliberately misused with inappropriate step sizes. For completeness,
      formulas describing a relationship between Jacobian accuracy and
      convergence rate order are included. 
  \item
      It is the authors' opinion that the above revisions offer the discussion of relative problem solving
      performance of the methods that the reviewer has requested, including consideration
      of speed and accuracy in solving real, non-linear problems.
  \end{itemize}}
  \end{enumerate}

  \subsection*{Smaller issues}
  \begin{enumerate}

  \item
	Pg. 2: In line 14, explain what does the term “algorithmically consistent” mean.

{\color{red}
  \begin{itemize}
     \item

      A brief definition for the convenience of the reader was added in the
      manuscript. But, algorithmically consistent is a fairly common term in
      plasticity/mechanics. This same meaning is used for 'consistent' refering
      to 'tangent matrices' in \cite[perez2012consistent, perez2010]. It means
      consistent with integrating the plasticity model under the Khun-Tucker
      constraints. Integrating the plasticity model means, given a strain
      increment, determining a stress that does not exceed the yield stress,
      and by doing so computing a linear slope, called a consistent tangent
      modulus, that relates the strain increment to a correct stress increment.
      An alogorithmically consistent tangent stiffness matrix is then one that
      uses the consistent tangent modulii, making it a more accurate
      approximation of the stiffness of the system.

@@

  \end{itemize}}

  \item
	Pg. 2: In line 36, what do you mean by “discrimination”?

{\color{red}  
\begin{itemize}
     \item

      The word, 'discrimination' has been removed from the manuscript.
      'Discrimination' was meant to mean 'measuring differences and making
      judgements on relative usefulness'. Since this meaning is probably
      understood from the word 'comparison' alone, the word 'discrimination'
      has been removed from the manuscript for clarity.

@@

  \end{itemize}}

` \item	
	Pg. 2: In line 37, explain what do you mean by “in-situ instrumentation”.

{\color{red}  
\begin{itemize}
     \item

      The words, 'in-situ' have been removed from the manuscript. 'In-situ' was
      meant to suggest that timing measurements were taken from within and by
      the program being analyzed itself, rather than by an external program,
      but it has been removed for clarity.

@@

{\color{red}  
\begin{itemize}
     \item
      The correction is made in the manuscript.
  \end{itemize}}

  \item 
Pg. 4: You may consider writing Eq. (2) with the truncation error

{\color{red}  
\begin{itemize}
     \item
      The correction is made in the manuscript.
  \end{itemize}}

\item
Pg. 6: In lines 9-11, please add references to the “Poisson problem” and to the
“mathematical optimization”.

{\color{red}  
\begin{itemize}
     \item
         [TODO]
         No changes made. 
  \end{itemize}}

\item
Pg. 7: Should Eqs. (4), (5), and (6) use “$\approx$” instead of “$=$”?

{\color{red}  
\begin{itemize}
     \item
         [TODO]
         Manuscript corrected. These expressions are approximations.
  \end{itemize}}

\item
Pg. 7: In line 28, note that (cid:126)u is not the current conﬁguration, but the displacement
ﬁeld. Please revise the sentence.

{\color{red}  
\begin{itemize}
    \item
        [TODO]
        Manuscript corrected. 'u' has been removed from the sentence since while the equations are functions of displacement, 'u' is not a configuration. 
        \end{itemize}}

\item
Pg. 7: Please provide an expression for ${F^int}_i$ in the model used.

{\color{red}  
\begin{itemize}
    \item
        [TODO]
        Manuscript corrected. In the section 'Peridigm and peridynamics' a constitutive model for the LPS is shown. On the page that was originally page 7, the force density corresponding to this constitutive model is expressed in descrete form.
  \end{itemize}}

\item
Pg. 9: In line 26, please add the reference [12] below to “coupling to molecular
simulations [29].”

{\color{red}  
\begin{itemize}
    \item
        The manuscript has been corrected.
  \end{itemize}}

\item
Pg. 10: Please add an illustration of the block undergoing tension in Section 2.2,
for clarity. Also, clarify the displacement boundary conditions; are they applied in a
volumetric layer?

{\color{red}  
\begin{itemize}
    \item
        [TODO]
        An conceptual illustration of the block undergoing tensions has been included. Additionally is a close-up illustration and some exposition explaining how displacement boundary conditions are achieved in peridynamics.
  \end{itemize}}

\item
Pg. 10: In line 40, you mentioned a bandwidth of 7. This seems to be clear in 1D.
Does this hold in higher dimensions? Please clarify.

{\color{red}  
\begin{itemize}
    \item
        [TODO]
        Bandwidth was referring to connections between nodes rather than components and degrees of freedom. It would be more clear to say, a bandwidth of 21 for the 3D case. The manuscript has been corrected with this change.
  \end{itemize}}

    \item
        In lines 47-49, you say “The speciﬁc parameters ... can be found ... in the
        paper repository.” It seems a good idea to have the paper repository with ﬁles that a
        user can eventually use. However, the content of the paper should be independent
        of the repository. Please include a table with the parameters used. Similarly, in
        page 11 (lines 11-13).

{\color{red}  
\begin{itemize}
    \item
        [TODO]
        Correction made in the manuscript. The test parameters not already reported throughout the manuscript include: material density
        and loading rates and information relevant to loading rates.
  \end{itemize}}

  \item
    Pg. 14: In line 55, is there any conclusion you can draw from the power law index
    $1.0 * 10^-6$?

{\color{red}  
\begin{itemize}
     \item
      Manuscript updated. More significant than the power law index is that the methods have the same power law relationship with computation time at all, meaning that there is some weak evidence that the methods have equal algorithm complexity. 
  \end{itemize}}

  \item
    Pgs. 14-: Please clarify the value of the meshsize used on each of the experiments.

{\color{red}  
\begin{itemize}
     \item
         No correction made. Justification follows.
         %
         Mesh sizes from smallest to largest were 1000, 2000, 4000, 8000,
         16000, 32000, 1000000 peridynamic nodes, but this is not relevant to
         the study, since the experiments were about the fullness of the
         tangent stiffness matrix and mesh size only places bounds on the
         mimimum and maximum fullness of the tangent stiffness matrix but does
         not determine it.  Furthermore, other users should not take their
         meshes of similar node count and make comparisons to the work based on that, since
         differing connectivity between nodes could make a 2000 node mesh more
         computationally challenging than a 1000000 node mesh.
         %
         The most extreme, but illustrative example of this is a stretched rope shaped mesh versus a cannonball shaped mesh of nearly overlapping nodes.
         %
         Instead, a better measure of problem size for this study was number of nonzero jacobian matrix elements, which was reported for the experiments in the work. Additionally, in \emph{Peridigm}, only the nonzero elements are stored in memory at all. This measure is a function of model dimensions, node spacing and horizon and not usefully related to number of nodes unless the meshes are at maximum or minimum connectivity.
      \end{itemize}}

  \item
    Pg. 16: Would it be possible to rescale Fig. 2, so that the dependence of the time
    on the number of cores will look indeed linear?

{\color{red}  
\begin{itemize}
     \item
     Manuscript is in error and has been corrected. An accurate statement is that CPU time and number of cores are inversely related.
  \end{itemize}}

  \item
    Pg. 16: In lines 54-56, you say “It should be pointed out that compared to the (cid:96)2 -
    norms of any of the TS matrices by themselves, the magnitude of the accuracy
    metric ... is relatively small.” Please include in the manuscript the explicit informa-
    tion on the relative error. It seems that you already have such information in the
    paper repository.

{\color{red}  
\begin{itemize}
     \item
[TODO]
    The manuscript has been corrected. 
    %
    The Frobenius norm of the TS matrices for each of the experiments was around 150 thousand, while the discrepancy between the matrices produced by the different methods, using the same metric, was on the order of 10 E-7. The relative error is on the order of 10 E-14. 
    \end{itemize}}

  \item
    Pg. 17: Include the values of the probe distances h used for each method.

{\color{red}  
\begin{itemize}
     \item
[TODO]
    Correction made in the manuscript, values used are collected and the reference to the material explaining these values was made more clear.
    %
    The probe distances used in Peridigm are automatically determined with a
    heuristic algorithm and are not set by the user. The algorithm for this in
    \emph{Peridigm} is the used for another Sandia software project called
    \emph{Adagio}. The probe distances are 1.E-6 times the node spacing [ADAGIO
    REF] for FD and CD which is indended to prevent the onset of roundoff error.
    1.E-100 was hard-coded for CS into the CS method implementation, while AD had no probe distance.
  \end{itemize}}

  \item
    Pg. 17: In lines 53-54, you say “Accuracy trends matched those for the single core
    tests.” Indeed, comparing Figs. 3 and 4, the results for FD seem to be quite the
    same in serial and in parallel. For CS the accuracy seems higher in serial. For CD,
    the accuracy is also higher in serial and, in serial, it has a ﬂuctuating behavior. Can
    you comment on these observations?

{\color{red}  
\begin{itemize}
     \item
[TODO]
	The fluctuation seems to be a function of the amount of nonzero TS matrix elements relative to total TS matrix elements. Information supporting this has been added to the manuscript. This may be related to the fact that neighborhood radius in terms of number of Peridynamic nodes was kept constant and may indicate that connectivity in the mesh oscillates when the mesh is refined incrementally. 
  \end{itemize}}

  \item
    Pgs. 17-18: What are the units of K ? If F int in Eq. (4) etc. is internal force, then K
    should have units of force/length. However, in Figs. 3 and 4 you use the unit of
    MPa, which is force/length2 . Please clarify.

{\color{red}  
\begin{itemize}
     \item
[TODO]
    The manuscript has been corrected to clarify that in Peridynamics, there is only force density rather than internal force. The units of K are actually force / length4, and this is corrected in the manuscript.
  \end{itemize}}

    \item
        Pg. 19: Are the units of the vertical axis in Fig. 5 “seconds”, as in Figs. 1 and 2?
        Please indicate that. Also, is the label of the vertical axis in Fig. 6 correct?

{\color{red}  
\begin{itemize}
     \item
[TODO]
        The manuscript has been corrected. The labels on the vertical axis in Fig. 5 were incorrect. 
  \end{itemize}}

Pg. 20: In lines 32-35, clarify what “byte-copyable” mean and why is that important.

{\color{red}  
\begin{itemize}
     \item
     Clarification made in manuscript, new reference added. 
     %
     'Byte-copiable' means that members of an array of data elements are all of
     the same size in memory so that if one knew the position of the first
     element in the array, the position of all other elements in the array
     could be inferred with the additional information of the length of the
     array. This is important, in particular, to the designers and users of the
     thousands of MIC accelerator cards installed in the \emph{Stampede}
     computer system. MIC accelerator cards permit only the transfer of
     byte-copiable data from the host computer system for use with the '#pragma
     offload' mode, which is to say most code written for the MIC. This is even
     more important when it is realized that the great majority of
     \emph{Stampede}'s theoretical processing power, the figure used to advertise to potential clients, is counted from these MIC
     accelerator cards. Meaning that if an AD user wanted to use any of that
     power, they would be out of luck since not even the full time personel at
     TACC could manage to compile the AD library Trilinos::Sacado for the MIC.
     A more 'casual' computational scientist would find the task impossible.
  \end{itemize}}

 \item
    Pg. 21: Please include a section title for the references.

{\color{red}  
\begin{itemize}
     \item
         [TODO]
         Manuscript corrected.
  \end{itemize}}

 \item
    Pg. 24: In line 23, you may add the truncation error O(h2 ) to the expression. Oth-
    erwise, you may need to replace the “=” sign by the “≈” sign. However, adding
    the error term would make more clear the limiting expression when h → 0. Also,
    replace the comma by a period.

{\color{red}  
\begin{itemize}
     \item
         [TODO]
         Manuscript corrected.
  \end{itemize}}

\item    
Pg. 24: Why do we need S and X in Appendix B?

{\color{red}  
\begin{itemize}
     \item
     All the members of set X need to be members of set S otherwise the approximations don't work. X are numbers from the 'real' world and S are numbers from the computer world. For example: "http://en.wikipedia.org/wiki/IEEE_floating_point" shows that computers and languages following IEEE standards don't represent very large numbers with very fine precision at the same time due to memory constraints, and there are upper and lower bounds on the magnitudes of nonzero numbers that can be represented. Obviously, approximation error is expected, but the fact that a user doesn't actually get 'all' the numbers is something that is not obvious.
  \end{itemize}}

\item
    Pgs. 25-26: Are the functions in the second column of the table in 2(c) (page 25)
    approximations? Why do you label the column “Approximates Function”? Simi-
    larly, in line 16 (page 26) you say “approximated by the partial derivative”; is it an
    approximation?

{\color{red}  
\begin{itemize}
     \item
      A computer can only provide finite precision approximations, of approxmations, of functions. Approximations of derivatives of functions are also approximations, therefor the label is correct in both cases. For example: "https://sourceware.org/git/?p=glibc.git;a=blob;f=sysdeps/x86_64/fpu/e_expl.S" is the implementation of the $exp(x)$ function on 80-bit extended precision floating point numbers in the popular GCC compiler. This implementation is the most accurate that GCC provides for the $e$ function on real numbers, but ultimately is still a discrete numerical algorithm.
  \end{itemize}}

  \item
Pg. 26: Please clarify whether storage is needed in automatic differentiation. For
example, do we need to store w, v , and u?

{\color{red}  
\begin{itemize}
     \item
         [TODO]
      Clarification is added in the manuscript. For an actual AD system like Trilinos::Sacado or ADIFOR, w, v and u (stand-ins for simple or elementary math operations) need to be written in code differently from the built-in implementation to replace all math operations as part of a complete AD datatype. What doesn't absolutely have to be stored are the evaluations of w, v and u. Not storing the evaluations happens in 'forward' AD, storing the evaluations happens in 'backward' or 'reverse accumulation' AD.  
  \end{itemize}}

### 2014-03-02 jacobian_calc_CMAME.tex
added mesh diagrams, fixed spelling throughout

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing complex-step in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. This open source resource is available at the correposinding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
 or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

@@

For the reader's convenience the foward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left ( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left ( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
    Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

@@

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Frechet derivatives of matix to matrix functions
\cite{al2010complex}. To compute the Jacobian of the residual in the
peridynamics simulation as discussed in the instant work is analogous is
computing a Frechet derivative. Where the application of the CS derivatives in
the prior work is limited to a particular class of abstract functions to
determine the CS method's suitability for use in condition number estimation, the
instant work applies CS to solving a solid mechanics problem. The instant work is not
about computing Frechet derivatives with CS but moreover using them in a
numerical method to solve a set of equations describing a physical system.

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is 
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
Where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the magnitude of perturbation called the \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}. From the well-known 1-dimensional formulas for FD and CD, it is understood that the accuracy of these expressions is dependent theoretically upon selecting a small step-size $h$, however experiments shown in \cite{squire1998using} demonstrate that too small an $h$ can lead to inaccuracy from ``subtractive-cancellation''. Therefore it is fair to say that a general method for selecting $h$ in practice is unobtainable.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} K_{ij} = \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h
\hat{e}_j))}{h}, \end{equation}
%
where the internal force, $F^{int}$ is now treated as function of complex
vectors,  $\vec{u}$. The small perturbation, $h$, is now carried out on the
imaginary axis and only the real coefficient to the imaginary part of the
vector returned by $F^{int}$ is kept, hence the $\mbox{Im}(\cdot)$ operation.
As highlighted in Section~\ref{sec:CSmethod}, notice that there is no
subtraction operation taking place; therefore $h$ can be made very small to
yield highly accurate derivatives. It should be mentioned that using
complex-step to compute tangent-stiffness matrices was done in
\cite{perez2000numerical,perez2012numerical}, but those references did not
investigate the performance of complex-step in a massively parallel setting nor
make comparisons to AD.

When using automatic-differentiation to calculate a tangent-stiffness matrix,
one only needs to follow the definition of the tangent-stiffness matrix and
issue the correct commands to the AD system. The formula is the same as the
continuum formula
%
\begin{equation} K_{ij} = \frac{\partial F_i^{int}(\vec{u})}{\partial u_j},
\end{equation}
%
but, with the caveat that $F^{int}$ is evaluated in a way that is
algorithmically consistent, and therefore should yield quadratic convergence
with Newton's method.

@@

The type of material model used in this study is referred to as a linear
peridynamic solid, having what is called a state based constitutive model. For
the convenience of the reader, vector states will be introduced. Then the state
based peridynamic equation of motion will be shown, followed by a description
of the linear peridynamic solid model used in this work. After the peridynamic
equation of motion and a consitutive model have been discussed, a definition of
the tangent stiffness matrix is given. The following discussion is largely
taken from \cite{silling:psa} but reproduced here for clarity.

@@

\note[Michael]{re: I.2} \add[Michael]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simualtion.}
The tangent stiffness matrix is the Frechet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Frechet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

@@

\subsubsection{Implementing the complex-step method in Peridigm} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
\change[Michael]{STL}{Standard Template Library} vectors, are hard coded to be
of type {\tt double} and can not be simply be declared as having a complex data
type. \note[Michael]{re: I.3}\add[Michael]{For the reader's convenience,
    \emph{Epetra} vectors and STL vectors are what are known as 'container
    classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exihibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code.} Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \note[Michael]{re: I.3}\add[Michael]{For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types.} No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules.} The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists; therefore,
many advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic, and distributed data structures, \note[Michael]{re:
I.3}\add[Michael]{for example \emph{Epetra} vectors, were used in the basic
    code and by necessity in the modified code. These concepts are not vital to
understanding the complex step method iteself, but may be necessary to understand complex
step in \emph{Peridigm}.} To see CS and other methods in a more narrow context,
it is advised that readers also take a look at the simple examples provided on
the corresponding author's website.

>>>>

The aim of this paper is not only to \note[Michael]{re: I.1}\change[Michael]{introduce the new}{study the} complex-step method \change[Michael]{in
the context of}{of} evaluating tangent-stiffness matrices, but to serve as a review
for other differentiation techniques useful for solving non-linear systems with
Newton's method. After presenting background information on the underlying
methods to be discussed, presented in this paper are: In subsection~\ref{subsec:TS} a description of
tangent-stiffness matrices, detailed directions for producing
tangent-stiffness matrices with each of the methods identified above, and a
description and justification of the \note[Michael]{re: I.1}\remove[Michael]{new}complex-step method for calculating
tangent-stiffness matrices, in sub-subsection~\ref{subsubsec:Impl} a description of implementing complex-step in
the \emph{Peridigm} software, in section~\ref{JGAM} a description of the quantities-of-interest
used to rank the methods, in section~\ref{sec:Res} a presentation and analysis of the results of the
comparative study, and in section~\ref{sec:Conc} conclusions and thoughts on potential future work.
Finally, in the interest of replicable research, the reader is referred to the
corresponding author's website for \emph{Peridigm} C++ source-code and data
necessary to reproduce the results presented here. Additionally, included on
the website is an example serial C++ library with classes for solving
non-linear systems using the techniques discussed here, as well as two
validation and verification example problems which show how to use the library.
The purpose of the serial C++ library is to demonstrate in a less complex
manner than in the \emph{Peridigm} code how one would use
automatic-differentiation or complex-step to better encourage their use and
discussion. This open source resource is available at the corresponding author's website: http://idl.utsa.edu/jtfoster/paper-repositories/
 or directly at https://github.com/utsa-idl/Complex_Step_Jacobian_Project_Archive.

@@

For the reader's convenience the forward difference equivalent to~\ref{eqn:complexFirstDeriv} is
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x)}{h} +\mathcal{O}\left ( h \right). \label{eqn:forwardFirstDeriv} \end{equation},
%
while the central difference equivalent to~\ref{eqn:complexFirstDeriv} is,
%
\begin{equation} \frac{\partial f( x )}{\partial x} = \frac{f (x + h) - f (x - h)}{2h} +\mathcal{O}\left ( h^2 \right). \label{eqn:centeredFirstDeriv} \end{equation} \cite[Chap. 4.1.3]{chapra2010}.
%
    Note that the order of the truncation error in~\ref{eqn:forwardFirstDeriv} indicates that for a given $h$ forward difference is less accurate an approximation than central difference or complex step.

@@

A work by A. H. Al-Mohy and N. J. Higham investigates the use of complex step
in computing Frechet derivatives of matrix to matrix functions
\cite{al2010complex}. To compute the Jacobian of the residual in the
peridynamics simulation as discussed in the instant work is analogous is
computing a Frechet derivative. Where the application of the CS derivatives in
the prior work is limited to a particular class of abstract functions to
determine the CS method's suitability for use in condition number estimation, the
instant work applies CS to solving a solid mechanics problem. The instant work is not
about computing Frechet derivatives with CS but moreover using them in a
numerical method to solve a set of equations describing a physical system.

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is 
%
\begin{equation} 
  K_{ij} \approx \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%

Where $K_{ij}$ is the indicial notation representation of an element of the
tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in
the numerator is the internal force, $F_i^{int}$ evaluated in the current
deformed configuration, as a function of the displacement, $\vec{u}$,  plus a small perturbation $h$ in the
direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding
to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is
$F_i^{int}$ evaluated in the current configuration. The denominator is the
magnitude of perturbation called the \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} \approx \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}. From the well-known 1-dimensional formulas for FD and CD, it is understood that the accuracy of these expressions is dependent theoretically upon selecting a small step-size $h$, however experiments shown in \cite{squire1998using} demonstrate that too small an $h$ can lead to inaccuracy from ``subtractive-cancellation''. Therefore it is fair to say that a general method for selecting $h$ in practice is unobtainable.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} K_{ij} \approx \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h
\hat{e}_j))}{h}, \end{equation}
%
where the internal force, $F^{int}$ is now treated as function of complex
vectors,  $\vec{u}$. The small perturbation, $h$, is now carried out on the
imaginary axis and only the real coefficient to the imaginary part of the
vector returned by $F^{int}$ is kept, hence the $\mbox{Im}(\cdot)$ operation.
As highlighted in Section~\ref{sec:CSmethod}, notice that there is no
subtraction operation taking place; therefore $h$ can be made very small to
yield highly accurate derivatives. It should be mentioned that using
complex-step to compute tangent-stiffness matrices was done in
\cite{perez2000numerical,perez2012numerical}, but those references did not
investigate the performance of complex-step in a massively parallel setting nor
make comparisons to AD.

When using automatic-differentiation to calculate a tangent-stiffness matrix,
one only needs to follow the definition of the tangent-stiffness matrix and
issue the correct commands to the AD system. The formula is the same as the
continuum formula
%
\begin{equation} K_{ij} \approx \frac{\partial F_i^{int}(\vec{u})}{\partial u_j},
\end{equation}
%
but, with the caveat that $F^{int}$ is evaluated in a way that is
algorithmically consistent, and therefore should yield quadratic convergence
with Newton's method.

@@

The type of material model used in this study is referred to as a linear
peridynamic solid, having what is called a state based constitutive model. For
the convenience of the reader, vector states will be introduced. Then the state
based peridynamic equation of motion will be shown, followed by a description
of the linear peridynamic solid model used in this work. After the peridynamic
equation of motion and a constitutive model have been discussed, a definition of
the tangent stiffness matrix is given. The following discussion is largely
taken from \cite{silling:psa} but reproduced here for clarity.

@@

\note[Michael]{re: I.2} \add[Michael]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simulation.}
The tangent stiffness matrix is the Frechet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
static peridynamic momentum equation. The internal force
density is computed in the peridynamic momentum equation by computing the
volume integral over a horizon of the Frechet derivative of the 
strain energy density state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

@@

\subsubsection{Implementing the complex-step method in Peridigm} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++
\change[Michael]{STL}{Standard Template Library} vectors, are hard coded to be
of type {\tt double} and can not be simply be declared as having a complex data
type. \note[Michael]{re: I.3}\add[Michael]{For the reader's convenience,
    \emph{Epetra} vectors and STL vectors are what are known as 'container
    classes', whose purpose in object oriented computer programming is to
    automate the creation, accessing, modification and destruction of the
    underlying contained data. Templated container classes allow for user
    choice in what properties that underlying data exhibits.  For example, in
    this work, complex arithmetic cannot be performed on the data contained
    within an \emph{Epetra} vector because that data, of type {\tt double} does
    not have the property of following the rules of complex arithmetic. The
    reason the non-templated \emph{Epetra} vector could not be dispensed with
    for the purposes of this work is because the \emph{Epetra} vector was
tightly integrated with \emph{Peridigm's} parallel communication strategy and
linear algebra code.} Program methods that were used in the course of applying
the FD method were copied, renamed and re-written to follow the CS formula.
Because of restrictions imposed by the container class being used, this
involved changing the data type of intermediate variables which were locally
scoped, and dynamically casting persistent memory variables to complex data
types where necessary. \note[Michael]{re: I.3}\add[Michael]{For the reader's
convenience, 'locally scoped' in the C++ programming language refers to
data which is created within a context more specific than the context of
the entire program, or 'translation unit'.  Most data created this way is
destroyed when that specific context is left. The topic of contexts in C++
is very broad, so for the purposes of this work context refers to a program
method or function that begins, concludes and yields a result, also
destroying locally scoped data. Dynamic casting for the purposes of this
work refers to copying the value of data contained within an \emph{Epetra}
vector into locally scoped data that has the property of following the rules of
complex arithmetic, where a mapping called 'type casting' allows the
translation of data between different types.} No attempt was made to ``tune''
the CS code in any way for performance. The algorithm  was implemented in such
as way that it was in one-to-one correspondence with the FD and CD methods;
however, it must be noted that the overhead associated with the dynamic casts
in the CS method likely hurts its overall performance and this issue could be
alleviated by using the next generation of templated \emph{TPetra} vectors
available in \emph{Trilinos} that are capable of taking an explicit
instantiation of any data type including {\tt complex}. \note[Michael]{re:
I.3}\add[Michael]{For the purposes of this work, an explicit instantiation
refers to requesting the creation of an object, or instance, of a templated
class or container class that holds data which has the property of
following desired mathematic or logic rules.} The version of \emph{Peridigm}
modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists; therefore,
many advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic, and distributed data structures, \note[Michael]{re:
I.3}\add[Michael]{for example \emph{Epetra} vectors, were used in the basic
    code and by necessity in the modified code. These concepts are not vital to
understanding the complex step method itself, but may be necessary to understand complex
step in \emph{Peridigm}.} To see CS and other methods in a more narrow context,
it is advised that readers also take a look at the simple examples provided on
the corresponding author's website.
